---
title: "Analysis of storms impact on population health and economy in US"
author: "Georgy Makarov"
date: 'April 21, 2020'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

Storms cause public health and economic problems in US. Preventing fatalities, 
injuries and asset loss are major concern for local authorities.

The goal of this analysis was to identify the most hazardous weather events in
terms of population health and economy in US. This analysis is based on US National
Oceanic and Atmospheric Administration data covering events from 1950 to 2011.

The analysis shows that the most harmful event for population health is tornado.
The most harmful event for economy is flood.

## Prerequisite packages

This analysis used four packages: *reshape2, dplyr, lubridate, ggplot2*.

```{r, warning=FALSE, message=FALSE}

library(reshape2)
library(dplyr)
library(lubridate)
library(ggplot2)

```

## Load data

Data from [NOAA]("http://www.noaa.gov") is available from cloud storage 
[here]("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"). 
Data description 
is [here]("http://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf").
More information about the storm is available in [FAQ]("http://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf").

The approach to loading the data is to download compressed file from URL if not 
found in working directory. Then load the data using *read.csv* and validate the
data by checking the file size and data dimensions, which are available in 
[supporting forum]("http://www.coursera.org/learn/reproducible-research/discussions/weeks/4/threads/38y35MMiEeiERhLphT2-QA").

```{r, warning=FALSE, message=FALSE, cache=TRUE}

raw_data_url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if (!file.exists("repdata_data_StormData.csv.bz2")) {
    download.file(url = raw_data_url, 
                  destfile = "repdata_data_StormData.csv.bz2")
    }

file_info <- file.info("repdata_data_StormData.csv.bz2")
storm_data <- read.csv("repdata_data_StormData.csv.bz2", sep = ",", header = TRUE)

stopifnot(file_info[,1] == 49177144)
stopifnot(dim(storm_data) == c(902297,37))

```

## Data processing

### Subsetting dataset

There are seven total variables required to perform this analysis: event type,
fatalities, injuries, damage to property, property damage multiplier, crop 
damage, crop damage multiplier. Those can be obtained from the names of the 
variables in the raw data.

```{r}

names(storm_data)

```

The list of necessary variables follows:  
- EVTYPE - event type;  
- FATALITIES - number of fatalities in event;  
- INJURIES - number of injuries in event;  
- PROPDMG - property damage;  
- PROPDMGEXP - property damage multiplier;  
- CROPDMG - crop damage;  
- CROPDMGEXP - crop damage multiplier.

The dataset is reduced to these variables.

```{r}

subset_data1 <- storm_data %>% 
    select(event_type = EVTYPE,
           fatalities = FATALITIES,
           injuries = INJURIES,
           property_damage = PROPDMG,
           property_multiplier = PROPDMGEXP,
           crop_damage = CROPDMG,
           crop_multiplier = CROPDMGEXP)

```

Part of observations contains event type *NONE* - we kick those observations.
Additionally, we only need observations with values greater than zero.

```{r}

subset_data2 <- subset_data1 %>% 
    filter(fatalities >0 | injuries > 0 | 
               property_damage > 0 | crop_damage > 0 & event_type != "NONE")

```

The subset has reduced to 254633 rows. Check that there are no missiong values 
in the subset.

```{r}

dim(subset_data2)
sum(complete.cases(subset_data2))

```

### Modifying event type

The official events type are 48. The subset contains 488 unique events type.

```{r}

length(unique(subset_data2$event_type))

```

The reason behind large number of types is that the dataset contains typos and
similar events in capital letters like: *wind* and *WIND*. 

```{r}

unique(subset_data2$event_type)

```


One way to solve this problem is to convert all event types to lowercase 
characters. This action reduced the number of event types to 447.

```{r}

subset_data3 <- subset_data2 %>% 
    mutate(event_type = tolower(event_type))
length(unique(subset_data3$event_type))

```

Further reduction of the dataset is acheived by grouping the events type using
the key word from the official event type names. The events that do not match 
any key word are grouped into *other*.

```{r}

subset_data3$event <- "other"
subset_data3$event[grep("avalanche", subset_data3$event_type)] <- "avalanche"
subset_data3$event[grep("blizzard", subset_data3$event_type)] <- "snow"
subset_data3$event[grep("flood", subset_data3$event_type)] <- "flood"
subset_data3$event[grep("wind", subset_data3$event_type)] <- "wind"
subset_data3$event[grep("fog", subset_data3$event_type)] <- "fog"
subset_data3$event[grep("cold", subset_data3$event_type)] <- "cold"
subset_data3$event[grep("chill", subset_data3$event_type)] <- "cold"
subset_data3$event[grep("frost", subset_data3$event_type)] <- "cold"
subset_data3$event[grep("freeze", subset_data3$event_type)] <- "cold"
subset_data3$event[grep("tornado", subset_data3$event_type)] <- "tornado"
subset_data3$event[grep("hail", subset_data3$event_type)] <- "hail"
subset_data3$event[grep("winds", subset_data3$event_type)] <- "wind"
subset_data3$event[grep("win", subset_data3$event_type)] <- "wind"
subset_data3$event[grep("wins", subset_data3$event_type)] <- "wind"
subset_data3$event[grep("storm", subset_data3$event_type)] <- "storm"
subset_data3$event[grep("rainstorm", subset_data3$event_type)] <- "storm"
subset_data3$event[grep("thunderstorm", subset_data3$event_type)] <- "storm"
subset_data3$event[grep("snow", subset_data3$event_type)] <- "snow"
subset_data3$event[grep("rain", subset_data3$event_type)] <- "rain"
subset_data3$event[grep("heat", subset_data3$event_type)] <- "heat"
subset_data3$event[grep("hurricane", subset_data3$event_type)] <- "hurricane"
subset_data3$event[grep("fld", subset_data3$event_type)] <- "flood"
subset_data3$event[grep("current", subset_data3$event_type)] <- "waves"
subset_data3$event[grep("surf", subset_data3$event_type)] <- "waves"
subset_data3$event[grep("fire", subset_data3$event_type)] <- "fire"
subset_data3$event[grep("water", subset_data3$event_type)] <- "flood"
subset_data3$event[grep("wave", subset_data3$event_type)] <- "waves"
subset_data3$event[grep("tsunami", subset_data3$event_type)] <- "waves"
subset_data3$event[grep("lightning", subset_data3$event_type)] <- "storm"
subset_data3$event[grep("warm", subset_data3$event_type)] <- "heat"
subset_data3$event[grep("torndao", subset_data3$event_type)] <- "tornado"
subset_data3$event[grep("high tides", subset_data3$event_type)] <- "waves"
subset_data3$event[grep("high seas", subset_data3$event_type)] <- "waves"
subset_data3$event[grep("marine mishap", subset_data3$event_type)] <- "waves"
subset_data3$event[grep("slide", subset_data3$event_type)] <- "slides"
subset_data3$event[grep("dust devil", subset_data3$event_type)] <- "heat"
subset_data3$event[grep("dry microburst", subset_data3$event_type)] <- "heat"
subset_data3$event[grep("low temperature", subset_data3$event_type)] <- "cold"
subset_data3$event[grep("freezing spray", subset_data3$event_type)] <- "cold"
subset_data3$event[grep("dam break", subset_data3$event_type)] <- "flood"

```

This operation has reduced the dataset events type to 16 unique values.

```{r}

length(unique(subset_data3$event))

```

Top-10 events type by the number of rows cover 99% of all observations.

```{r}

sum(sort(table(subset_data3$event), decreasing = TRUE)[1:10]) / nrow(subset_data3)

```

The dataset is reduced to top-10 events.

```{r}

sum(sort(table(subset_data3$event), decreasing = TRUE)[1:10])
sort(table(subset_data3$event), decreasing = TRUE)[1:10]
subset_data4 <- subset_data3 %>% 
    filter(event %in% c("wind", "storm", "tornado", "flood", "hail", "snow", 
                        "fire", "rain", "heat", "waves"))
dim(subset_data4)

```

### Cleaning units

Property damage multiplier and crop damage multiplier contain different units.

```{r}

unique(subset_data4$property_multiplier)
unique(subset_data4$crop_multiplier)

```


Some units like k or K, m or M, B reflect thousands, millions and billions of USD.
Numbers from 1 to 8 show the power coefficient for the multiplier like: $$ e = 10^k $$

where $ e $ is multiplier and $ k $ is from 1 to 8.

Other symbols do not transform to currency directly. For the purpose of this
analysis those symbols are condidered to be 1 USD.

```{r}

subset_data5 <- subset_data4 %>% 
    mutate(property_multiplier = tolower(property_multiplier),
           crop_multiplier = tolower(crop_multiplier))

subset_data5$property_multiplier <- as.character(subset_data5$property_multiplier)
subset_data5$property_multiplier[is.na(subset_data5$property_multiplier)] <- 0
subset_data5$property_multiplier[!grepl("k|m|b|h|2|3|4|5|6|7", subset_data5$property_multiplier)] <- 0
subset_data5$property_multiplier[grep("k", subset_data5$property_multiplier)] <- "3"
subset_data5$property_multiplier[grep("m", subset_data5$property_multiplier)] <- "6"
subset_data5$property_multiplier[grep("b", subset_data5$property_multiplier)] <- "9"
subset_data5$property_multiplier[grep("h", subset_data5$property_multiplier)] <- "2"

subset_data5$crop_multiplier <- as.character(subset_data5$crop_multiplier)
subset_data5$crop_multiplier[is.na(subset_data5$crop_multiplier)] <- 0
subset_data5$crop_multiplier[!grepl("k|m|b|h|2|3|4|5|6|7", subset_data5$crop_multiplier)] <- 0
subset_data5$crop_multiplier[grep("k", subset_data5$crop_multiplier)] <- "3"
subset_data5$crop_multiplier[grep("m", subset_data5$crop_multiplier)] <- "6"
subset_data5$crop_multiplier[grep("b", subset_data5$crop_multiplier)] <- "9"
subset_data5$crop_multiplier[grep("h", subset_data5$crop_multiplier)] <- "2"

subset_data5$property_multiplier <- as.numeric(as.character(subset_data5$property_multiplier))
subset_data5$crop_multiplier <- as.numeric(as.character(subset_data5$crop_multiplier))

subset_data5 <- subset_data5 %>% 
    mutate(prop_damage = property_damage * 10^property_multiplier,
           crop_damage = crop_damage * 10^crop_multiplier) %>% 
    select(event, fatalities, injuries, crop_damage, prop_damage)

head(subset_data5)

```

### Data aggregation by health and economic results

The data is aggregated by the type of event using *summarise* function resulting
in 2 datasets: health and economic. Economic figures are transformed into 
billions.

```{r}

health_dt <- subset_data5 %>% 
    group_by(event) %>% summarise(fatalities = sum(fatalities),
                                  injuries = sum(injuries))
fatal_h <- health_dt %>% 
    select(event, result = fatalities) %>% 
    mutate(type = "fatal")
injuries_h <- health_dt %>% 
    select(event, result = injuries) %>% 
    mutate(type = "injury")
health_dt <- rbind(fatal_h, injuries_h)
health_dt

economic_dt <- subset_data5 %>% 
    group_by(event) %>% summarise(crop_damage_bln = sum(crop_damage) / 1000000000, 
                                  prop_damage_bln = sum(prop_damage) / 1000000000)
prop_econ <- economic_dt %>% 
    select(event, result = prop_damage_bln) %>% 
    mutate(type = "property damage")
crop_econ <- economic_dt %>% 
    select(event, result = crop_damage_bln) %>% 
    mutate(type = "crop damage")
economic_dt <- rbind(prop_econ, crop_econ)
economic_dt

```

## Results

### Which type of events are most harmful to population health

Tornado is the most harmful event of all severe weather conditions in US. Total
number of injuries and deaths is 96997, which is 7 times more than the storms.

```{r}

health_dts <- health_dt %>% 
    group_by(event) %>% summarise(result = sum(result)) %>% 
    arrange(desc(result))
health_dts

```

Distribution of health impact is given in the figure below.

```{r}

ggplot(health_dt) +
    theme_bw() +
    ggtitle("Weather impact on population health") +
    xlab("Event type") + ylab("Fatalities and Injuries") +
    geom_col(aes(reorder(event, result), result, fill = type))

```

### Which types of events have the greatest economic consequences

In economic terms floods caused 181 bln USD loss, which makes them the most
harmful event.

```{r}

economic_dts <- economic_dt %>% 
    group_by(event) %>% summarise(result = sum(result)) %>% 
    arrange(desc(result))
economic_dts

```

Distribution of economic impact is given in the figure below.

```{r}

ggplot(economic_dt) +
    theme_bw() +
    ggtitle("Weather impact on economy") +
    xlab("Event type") + ylab("Loss in bln USD") +
    geom_col(aes(reorder(event, result), result, fill = type))

```

## Conclusion

Tornado is the major cause of death or injury of other types of severe weather 
conditions. It caused almost 100k deaths and injuries. As for the economy, the
most harmful is the flood with 1814 bln USD in total loss. The major source of 
loss is property damage.

